Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-15 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Comments by Monticelli added and contributions updated

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work as only $\alpha$ is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the sign measurements of the order parameters see Ref.~\citenum{signpost}.

@@

In general there is a good agreement between the order parameters measured with different experimental NMR techniques: Almost all the 
reported values are within a variation of $\pm$0.02 (which is also the error estimate given by Gross et al.~\cite{gross97}) 
for all fully hydrated PC bilayer, regardless of the variation in their acyl chain composition and temperature.
Exceptions are the somewhat lower order parameters sometimes reported from measurements using $^1$H-$^{13}$C NMR~\cite{hong95a,hong95b,warschawski05}.
These experiments are not shown in Fig.~\ref{HGorderparameters} as the reported error bars are either relatively large~\cite{hong95a,hong95b}, 
or the spectral resolution is quite low and the numerical lineshape simulations have not been used in the analysis~\cite{warschawski05}.
Due to this end, it is highly likely that these reported lower order parameters are due to lower experimental 
accuracy and therefore we exclude them from our discussion. 
For more details, see Ref.~\citenum{accuracyPOST}. 
Motivated by the high experimental repeatability, we have highlighted in 
Fig.~\ref{HGorderparameters} the subjective sweet spots (light blue areas), within which we expect the calculated absolute 
values of the order parameters of a well-performing force field to fall.

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured 
(see Ref.~\citenum{accuracyPOST} for more discussion.) 
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

@@

\subsection{Author Contributions}
{\it Alexandru Botan} provided simulation results for charmm36-UA\\
{\it Fernando Favela} prepared, performed and analyzed simulations
with popc and cholesterol for the CHARMM36 All-Atom FF. \\
{\it Patrick F. J. Fuchs} Ran and analyzed the Poger simulations. Provided scientific information which significantly advanced the project (signs and forking of order parameters). \\
{\it Matti Javanainen} prepared and performed simulations with multiple lipid models and analyzed the results. Supervised the work of JT.\\
{\it Matej Kanduc} provided simulation results for the Berger DLPC model \\
{\it Waldemar Kulig} prepared the MD simulations with cholesterol for the MacRog FF \\
{\it Antti Lamberg} Prepared and performed simulations with Berger and its variants to show the importance of the signs and stereospecific labeling of the order parameters. \\
{\it Claire Loison} provided simulation results for charmm36-UA. \\
{\it Alexander Lyubartsev} provided simulation results for Högberg08 force field.\\
{\it Markus S. Miettinen} \todo{Please write a text summarizing your contribution} \\
{\it Luca Monticelli} Critical discussions in all phases of the project. Collaboration with Jukka  M{\"a}{\"a}tt{\"a}. \\
{\it Jukka M{\"a}{\"a}tt{\"a}}  prepared and performed simulations with Berger and Slipids models and analyzed the results.\\
{\it O. H. Samuli Ollila} Designed and managed the work. Ran and analyzed several simulations. Wrote the manuscript.  \\
{\it Marius Retegan} Prepared and validated the GROMACS compatible parameter files for GAFFlipid and Lipid14 force fields.\\
{\it Tomasz Rog} Provided the topologies for the MacRog model and also the full hydration simulation data with this model. \\
{\it Hubert Santuz} prepared and performed the cholesterol simulations with CHARMM36 and analyzed the results. 
Made significant contribution in the data management of the project.\\
{\it Joona Tynkkynen} prepared and performed the dehydration simulations with the MacRog FF.\\

>>>>

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work, except for $\alpha$ which is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the determination of the sign of the order parameters see Ref.~\citenum{signpost}.

@@

In general there is a good agreement between the order parameters measured with different experimental NMR techniques: Almost all the 
reported values are within a variation of $\pm$0.02 (which is also the error estimate given by Gross et al.~\cite{gross97}) 
for all fully hydrated PC bilayer, regardless of the variation in their acyl chain composition and temperature.
Exceptions are the somewhat lower order parameters sometimes reported from measurements using $^1$H-$^{13}$C NMR~\cite{hong95a,hong95b,warschawski05}.
These experiments are not shown in Fig.~\ref{HGorderparameters} as the reported error bars are either relatively large~\cite{hong95a,hong95b}, 
or the spectral resolution is quite low and the numerical lineshape simulations have not been used in the analysis~\cite{warschawski05}.
Due to this end, it is highly likely that these reported lower order parameters are due to lower experimental 
accuracy and therefore we exclude them from our discussion. 
For more details, see Ref.~\citenum{accuracyPOST}. 
Motivated by the high experimental reproducibility, we have highlighted in 
Fig.~\ref{HGorderparameters} the subjective sweet spots (light blue areas), within which we expect the calculated absolute 
values of the order parameters of a well-performing force field to fall.

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured 
(see Ref.~\citenum{accuracyPOST} for more discussion.) 
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

@@

\subsection{Author Contributions}
{\it Alexandru Botan} provided simulation results for charmm36-UA\\
{\it Fernando Favela} prepared, performed and analyzed simulations
with popc and cholesterol for the CHARMM36 All-Atom FF. \\
{\it Patrick F. J. Fuchs} Ran and analyzed the Poger simulations. Provided scientific information which significantly advanced the project (signs and forking of order parameters). \\
{\it Matti Javanainen} prepared and performed simulations with multiple lipid models and analyzed the results. Supervised the work of JT.\\
{\it Matej Kanduc} provided simulation results for the Berger DLPC model \\
{\it Waldemar Kulig} prepared the MD simulations with cholesterol for the MacRog FF \\
{\it Antti Lamberg} Prepared and performed simulations with Berger and its variants to show the importance of the signs and stereospecific labeling of the order parameters. \\
{\it Claire Loison} provided simulation results for charmm36-UA. \\
{\it Alexander Lyubartsev} provided simulation results for Högberg08 force field.\\
{\it Markus S. Miettinen} Co-designed the project with OHSO and supported in the work management. Provided trajectories for Berger DMPC; prepared parameter files for 72 lipids \\
{\it Luca Monticelli} Critical discussions in all phases of the project. Collaboration with Jukka  M{\"a}{\"a}tt{\"a}. \\
{\it Jukka M{\"a}{\"a}tt{\"a}}  prepared and performed simulations with Berger and Slipids models and analyzed the results.\\
{\it O. H. Samuli Ollila}Co-designed the project with MSM and managed the work. Ran and analyzed several simulations. Wrote the manuscript. \\
{\it Marius Retegan} Prepared and validated the GROMACS compatible parameter files for GAFFlipid and Lipid14 force fields.\\
{\it Tomasz Rog} Provided the topologies for the MacRog model and also the full hydration simulation data with this model. \\
{\it Hubert Santuz} prepared and performed the cholesterol simulations with CHARMM36 and analyzed the results. 
Made significant contribution in the data management of the project.\\
{\it Joona Tynkkynen} prepared and performed the dehydration simulations with the MacRog FF.\\
